Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-03-29 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Atmospheric concentrations of CO~2~ and water vapor can also vary across the vertical gradient.
Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. 
Vertical CO~2~ concentration gradients can influence the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1). 
Particularly, understory seedlings benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

@@

Vertical gradients in air temperature ($T_{air}$) across forest profiles are variable (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), being dependent upon factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019]. 
It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->]. 
Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021]. 
Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers [e.g., OSBS, Fig. `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, <!--@hadleyInfluenceKrummholzMat1987--> @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015-->; @mullerEvidenceEfficientNonevaporative2021].
These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.

>>>>

Atmospheric concentrations of CO~2~ and water vapor can also vary vertically.
Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. 
Vertical CO~2~ concentration gradients can drive gradients in the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and in rates of refixation of respired CO~2~ vertically, both of which tend to be higher at lower heights (Table 1). 
Particularly, understory seedlings benefit from elevated CO~2~ by having higher rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations concentrations likely have little effect on energy balance and metabolism across the vertical gradient.

@@

Vertical gradients in air temperature ($T_{air}$) differ widely (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`) and depend on factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019]. 
Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->]. 
Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021]. 
Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$, due to turbulent air mixing and interception of thermal radiative flux from the soil and the canopy by lower-canopy layers [e.g., OSBS, Fig. `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b; <!--@hadleyInfluenceKrummholzMat1987--> @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015--> @mullerEvidenceEfficientNonevaporative2021].
These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.
